Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-05-13 SIR_manuscript.tex
Old Fady's stuff are commented out for now

% %%%%%%%
\subsection{rate of returning tests} \label{app:omega}
{\bf The third goal} is to explore how changes in the rate of test return $\omega$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder[\Rnum]{\omega}$, which specifically depends on $\pder[\Delta]{\omega}$. We use
the linearization of $\Rnum$ around $\rho=0$ to show that there a non-monotonic relationship between $\Rnum$  and $\omega$. The taylor expansion of $\Delta$ at $\rho=0$ is
\begin{equation}
\label{eq:lin}
\Delta = \frac{\rho}{\omega\,\gamma(\omega+\gamma)} \Big(
\frac{w_I}{w_S}\,\omega^2\,p_I\,\theta\_c+(\frac{w_I}{w_S}+1)\,\gamma\omega\theta\_w\,+\gamma^2\,\theta\_w(2-\theta\_w)
\Big). 
\end{equation}
This results in
\begin{equation}
\label{eq:dlindo}
\pder[\Delta]{\omega} = \frac{\rho}{\omega^2\,(\omega+\gamma)^2} \Big(
(p_I\,\frac{w_I}{w_S}\theta\_c\,-(1+\frac{w_I}{w_S})\theta\_w)\omega^2+2\theta_w\,\gamma(\theta_w-2)\,\omega+\theta_w\,\gamma^2\,(\theta_w-2)
\Big),
\end{equation}
around $\rho=0$.
\ali{I stoped here!}

Perhaps counter-intuitively, the equation above does not predict that $\Rnum$ is monotone decreasing with respect to $\omega$. In other words; our model does not predict that returning test results more rapidly \textit{always} lower $\Rnum$. In order to gain insight into this intriguing behavior, we examine the zeroes of $\frac{\partial{\Rnum}}{\partial{\omega}}(\omega)$.
Defining the following quantity, $Q$, will help us write the roots of $\partial{\Rnum}/\partial{\omega}$ neatly as follows.
\begin{align}\label{eq:defQ}
    Q =& \frac{w_I}{w_S}\left(1-\frac{n_t-1}{n_w-1}p_I \right).
\end{align}
With that in mind, we can write the roots of $\partial{\Rnum}/\partial{\omega}$ as
\begin{align}
    \omega_1 =& \frac{\gamma}{-\sqrt{1-Q}-1} \\
    \omega_2 =& \frac{\gamma}{\sqrt{1-Q}-1}.
\end{align}

Note that the zeroes are real if and only if $Q < 1$. Note that have $\theta\_c > \theta\_w$, so if $p_I \approx 1$, we will have $Q < 0 < 1$. Thus, if we assume near-perfect test sensitivity, $\omega_1$ and $\omega_2$ will be real.

Assuming $\omega_1, \omega_2$ are real, it is easy to confirm that $\omega_1 < 0$ by looking at the denominator. To see that $\omega_2 > 0$, recall that $Q < 0$, so $\sqrt{1-Q} > 1$ and so $\sqrt{1-Q} -1 > 0$. Knowing that $\omega_1 < 0$, the only root of interest (i.e., biologically relevant quantity) is $\omega_2$.

We can prove that $\partial{\Rnum}/\partial{\omega} > 0$ when $\omega \in (0,\omega_2)$ and $\partial{\Rnum}/\partial{\omega} < 0$ when $\omega \in (\omega_2,\infty)$ by computing the limits of $\partial{\Rnum}/\partial{\omega}$ at $0$ and $\infty$ respectively. So it follows that $\Rnum$ has a global maximum with respect to $\omega$ at $\omega = \omega_2$.

Now we want to characterize the parameter regions on which $\partial{\Rnum}/\partial{\omega} < 0$ (i.e., the conditions under which returning test results more rapidly is favorable). By the previous analysis, this is equivalent to solving for $\omega > \omega_2$. So
\begin{align}
    &\omega > \omega_2 \nonumber \\
    &\omega > \frac{\gamma}{\sqrt{1-Q}-1} \nonumber \\
    &\sqrt{1-Q} > \frac{\gamma}{\omega}+1 \\
    &1-Q > (\frac{\gamma}{\omega}+1)^2.
\end{align}
Substituting in $Q$ from \eqref{eq:defQ} we have
\begin{align}
    &1-\frac{w_I}{w_S}\left(1-\frac{n_{t}-1}{n_{\omega}-1}P_{i}\right)> \left(\frac{\gamma}{\omega}+1\right)^2 \\
    &-\frac{w_I}{w_S}\left(1-\frac{n_{t}-1}{n_{\omega}-1}P_{i}\right)> \left(\frac{\gamma}{\omega}+1\right)^2+1 \\    
    &\frac{w_I}{w_S}\left(\frac{1-n_{t}}{1-n_{\omega}}P_{i}-1\right)> \left(\frac{\gamma}{\omega}+1\right)^2+1\label{eq:necsuf}.
\end{align}
Since all steps in deriving \eqref{eq:necsuf} are reversible, \eqref{eq:necsuf} gives a necessary and sufficient condition for $\omega > \omega_2$, which characterizes when returning tests more rapidly would cause a decrease in $\Rnum$.

@@

% %%%%%%%
\subsection{Expensive vs.\ cheap tests}

The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the \covid pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. In the analysis below, we compare the $\Rnum$ predicted by our model depending on the testing strategy.

Consider a test that allows us to test at rate $\rho_1$ and has sensitivity $P_{i,1}$, and another test that allows us to test at $\rho_2$ and has sensitivity $P_{i,2}$. Suppose that $\rho_1 > \rho_2$. Recall that the linearization of $\Rnum$ around $\rho \approx 0$ is given by $$\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 w_S} \Big(\gamma(-\theta\_w)(\gamma w_S+\omega w_I) + (-\theta\_c)p_I w_I \omega^2 \Big).$$

Treating $\Rnum$ as a function of $\rho$ and $P_i$,we can reduce the inequality $$\Rnum(\rho_2, p_{I,2}) < \Rnum(\rho_1, p_{I,1})$$ into

\begin{align}\label{eq:rho1vsrho2}
    &\rho_1\left(\gamma(-\theta\_w)(\gamma w_S + \omega w_I) + (-\theta\_c)p_{I,1}w_I\omega^2\right) - \rho_2\left(\gamma(-\theta\_w)(\gamma w_S + \omega w_I) + (-\theta\_c)p_{I, 2}w_I\omega^2\right) > 0 \nonumber \\
    &\vdots \nonumber \\
    &\frac{\rho_2P_{i, 2}-\rho_1P_{i, 1} }{\rho_1-\rho_2} > \frac{\theta\_w}{\theta\_c}\cdot \frac{\gamma(\gamma w_S + \omega w_I)}{\omega^2 w_I}
\end{align}

Note that the RHS is positive, thus a necessary condition for the inequality above to hold is that $\rho_2P_{i,2} > \rho_1P_{i,1}$, equivalently

\begin{equation}
\frac{P_{i,2}}{P_{i,1}} > \frac{\rho_1}{\rho_2}.
\end{equation}

To state an example of this, if test $A$ is three times as expensive as test $B$ (and hence one can test three times as many people with test $B$), using test $A$ rather than $B$ will be favorable only if test $A$ is at least 3 times more sensitive than test $B$. Note that this is a necessary but not sufficient condition, so even if test $A$ is three times more sensitive, it is still possible for test $B$ to be more effective.

\cref{eq:rho1vsrho2} tells us precisely when a test corresponding to $\rho_2, P_{i,2}$ will yield a lower $\Rnum$ than a test corresponding to $\rho_1, P_{i,1}$, where $\rho_1 > \rho_2$. Some of the qualitative \textit{trends} that favor test 2 (the higher-sensitivity test) include

\begin{itemize}
    \item individuals who test positive self-isolate much more than individuals who are waiting for their test result.
    \item the time it takes to return tests is much shorter than the mean infectious period.
    \item the testing intensity is much greater for infected individuals than susceptible individuals.
\end{itemize}

@@

We define $w\_{IS}=\frac{w_I}{w_S}$.
\begin{align}
\pder[\Delta]{w\_{IS}}= \dfrac{(\omega-\rho)\,(\omega(\omega-\rho\theta\_w)+\gamma(\omega-\rho))\,(\theta\_w\gamma+\theta\_c\omega\,p\_I)}
{(-\omega^2\,\gamma+\omega\,\gamma\,\rho-\gamma\,\rho\,\omega\,w\_{IS}-
\omega\,\gamma^2+\gamma^2\,\rho-\omega^2\,p_I\,\rho\,w\_{IS})^2},
\end{align}
which is a positive quantity. Thus, $\pder[\Rnum]{w\_{IS}}\leq 0$. Therefore, increasing the focus of testing on the infectious people will result in less transmission.  
\end{document}

>>>>

% %%%%%%%
\subsection{rate of returning tests} \label{app:omega}
{\bf The third goal} is to explore how changes in the rate of test return $\omega$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder[\Rnum]{\omega}$, which specifically depends on $\pder[\Delta]{\omega}$. We use
the linearization of $\Rnum$ around $\rho=0$ to show that there a non-monotonic relationship between $\Rnum$  and $\omega$. The taylor expansion of $\Delta$ at $\rho=0$ is
\begin{equation}
\label{eq:lin}
\Delta = \frac{\rho}{\omega\,\gamma(\omega+\gamma)} \Big(
\frac{w_I}{w_S}\,\omega^2\,p_I\,\theta\_c+(\frac{w_I}{w_S}+1)\,\gamma\omega\theta\_w\,+\gamma^2\,\theta\_w(2-\theta\_w)
\Big). 
\end{equation}
This results in
\begin{equation}
\label{eq:dlindo}
\pder[\Delta]{\omega} = \frac{\rho}{\omega^2\,(\omega+\gamma)^2} \Big(
(p_I\,\frac{w_I}{w_S}\theta\_c\,-(1+\frac{w_I}{w_S})\theta\_w)\omega^2+2\theta_w\,\gamma(\theta_w-2)\,\omega+\theta_w\,\gamma^2\,(\theta_w-2)
\Big),
\end{equation}
around $\rho=0$.
% \ali{I stoped here!}

@@

@@

We define $w\_{IS}=\frac{w_I}{w_S}$.
\begin{align}
\pder[\Delta]{w\_{IS}}= \dfrac{(\omega-\rho)\,(\omega(\omega-\rho\theta\_w)+\gamma(\omega-\rho))\,(\theta\_w\gamma+\theta\_c\omega\,p\_I)}
{(-\omega^2\,\gamma+\omega\,\gamma\,\rho-\gamma\,\rho\,\omega\,w\_{IS}-
\omega\,\gamma^2+\gamma^2\,\rho-\omega^2\,p_I\,\rho\,w\_{IS})^2},
\end{align}
which is a positive quantity. Thus, $\pder[\Rnum]{w\_{IS}}\leq 0$. Therefore, increasing the focus of testing on the infectious people will result in less transmission.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \todo{Fady:To be change accordingly please.} 
% Perhaps counter-intuitively, the equation above does not predict that $\Rnum$ is monotone decreasing with respect to $\omega$. In other words; our model does not predict that returning test results more rapidly \textit{always} lower $\Rnum$. In order to gain insight into this intriguing behavior, we examine the zeroes of $\frac{\partial{\Rnum}}{\partial{\omega}}(\omega)$.
% Defining the following quantity, $Q$, will help us write the roots of $\partial{\Rnum}/\partial{\omega}$ neatly as follows.
% \begin{align}\label{eq:defQ}
%     Q =& \frac{w_I}{w_S}\left(1-\frac{n_t-1}{n_w-1}p_I \right).
% \end{align}
% With that in mind, we can write the roots of $\partial{\Rnum}/\partial{\omega}$ as
% \begin{align}
%     \omega_1 =& \frac{\gamma}{-\sqrt{1-Q}-1} \\
%     \omega_2 =& \frac{\gamma}{\sqrt{1-Q}-1}.
% \end{align}
% 
% Note that the zeroes are real if and only if $Q < 1$. Note that have $\theta\_c > \theta\_w$, so if $p_I \approx 1$, we will have $Q < 0 < 1$. Thus, if we assume near-perfect test sensitivity, $\omega_1$ and $\omega_2$ will be real. 
% 
% Assuming $\omega_1, \omega_2$ are real, it is easy to confirm that $\omega_1 < 0$ by looking at the denominator. To see that $\omega_2 > 0$, recall that $Q < 0$, so $\sqrt{1-Q} > 1$ and so $\sqrt{1-Q} -1 > 0$. Knowing that $\omega_1 < 0$, the only root of interest (i.e., biologically relevant quantity) is $\omega_2$. 
% 
% We can prove that $\partial{\Rnum}/\partial{\omega} > 0$ when $\omega \in (0,\omega_2)$ and $\partial{\Rnum}/\partial{\omega} < 0$ when $\omega \in (\omega_2,\infty)$ by computing the limits of $\partial{\Rnum}/\partial{\omega}$ at $0$ and $\infty$ respectively. So it follows that $\Rnum$ has a global maximum with respect to $\omega$ at $\omega = \omega_2$.
% 
% Now we want to characterize the parameter regions on which $\partial{\Rnum}/\partial{\omega} < 0$ (i.e., the conditions under which returning test results more rapidly is favorable). By the previous analysis, this is equivalent to solving for $\omega > \omega_2$. So
% \begin{align}
%     &\omega > \omega_2 \nonumber \\
%     &\omega > \frac{\gamma}{\sqrt{1-Q}-1} \nonumber \\
%     &\sqrt{1-Q} > \frac{\gamma}{\omega}+1 \\
%     &1-Q > (\frac{\gamma}{\omega}+1)^2.
% \end{align}
% Substituting in $Q$ from \eqref{eq:defQ} we have
% \begin{align}
%     &1-\frac{w_I}{w_S}\left(1-\frac{n_{t}-1}{n_{\omega}-1}P_{i}\right)> \left(\frac{\gamma}{\omega}+1\right)^2 \\
%     &-\frac{w_I}{w_S}\left(1-\frac{n_{t}-1}{n_{\omega}-1}P_{i}\right)> \left(\frac{\gamma}{\omega}+1\right)^2+1 \\    
%     &\frac{w_I}{w_S}\left(\frac{1-n_{t}}{1-n_{\omega}}P_{i}-1\right)> \left(\frac{\gamma}{\omega}+1\right)^2+1\label{eq:necsuf}.
% \end{align}
% Since all steps in deriving \eqref{eq:necsuf} are reversible, \eqref{eq:necsuf} gives a necessary and sufficient condition for $\omega > \omega_2$, which characterizes when returning tests more rapidly would cause a decrease in $\Rnum$.
% % %%%%%%%
% \subsection{Expensive vs.\ cheap tests}
% 
% The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the \covid pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. In the analysis below, we compare the $\Rnum$ predicted by our model depending on the testing strategy. 
% 
% Consider a test that allows us to test at rate $\rho_1$ and has sensitivity $P_{i,1}$, and another test that allows us to test at $\rho_2$ and has sensitivity $P_{i,2}$. Suppose that $\rho_1 > \rho_2$. Recall that the linearization of $\Rnum$ around $\rho \approx 0$ is given by $$\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 w_S} \Big(\gamma(-\theta\_w)(\gamma w_S+\omega w_I) + (-\theta\_c)p_I w_I \omega^2 \Big).$$
% 
% 
% Treating $\Rnum$ as a function of $\rho$ and $P_i$,we can reduce the inequality $$\Rnum(\rho_2, p_{I,2}) < \Rnum(\rho_1, p_{I,1})$$ into 
% 
% \begin{align}\label{eq:rho1vsrho2}
%     &\rho_1\left(\gamma(-\theta\_w)(\gamma w_S + \omega w_I) + (-\theta\_c)p_{I,1}w_I\omega^2\right) - \rho_2\left(\gamma(-\theta\_w)(\gamma w_S + \omega w_I) + (-\theta\_c)p_{I, 2}w_I\omega^2\right) > 0 \nonumber \\
%     &\vdots \nonumber \\
%     &\frac{\rho_2P_{i, 2}-\rho_1P_{i, 1} }{\rho_1-\rho_2} > \frac{\theta\_w}{\theta\_c}\cdot \frac{\gamma(\gamma w_S + \omega w_I)}{\omega^2 w_I}
% \end{align}
% 
% Note that the RHS is positive, thus a necessary condition for the inequality above to hold is that $\rho_2P_{i,2} > \rho_1P_{i,1}$, equivalently 
% 
% \begin{equation}
% \frac{P_{i,2}}{P_{i,1}} > \frac{\rho_1}{\rho_2}.
% \end{equation}
% 
% To state an example of this, if test $A$ is three times as expensive as test $B$ (and hence one can test three times as many people with test $B$), using test $A$ rather than $B$ will be favorable only if test $A$ is at least 3 times more sensitive than test $B$. Note that this is a necessary but not sufficient condition, so even if test $A$ is three times more sensitive, it is still possible for test $B$ to be more effective. 
% 
% \cref{eq:rho1vsrho2} tells us precisely when a test corresponding to $\rho_2, P_{i,2}$ will yield a lower $\Rnum$ than a test corresponding to $\rho_1, P_{i,1}$, where $\rho_1 > \rho_2$. Some of the qualitative \textit{trends} that favor test 2 (the higher-sensitivity test) include
% 
% \begin{itemize}
%     \item individuals who test positive self-isolate much more than individuals who are waiting for their test result.
%     \item the time it takes to return tests is much shorter than the mean infectious period.
%     \item the testing intensity is much greater for infected individuals than susceptible individuals.
% \end{itemize}

\end{document}
